---
title: "Movement Analysis"
author: "Nicola Noger"
format: html
editor: visual
engine: knitr
---

## Solution for Movement Analysis

In this document, I solve the tasks for Movement Analysis of the course Spatiotemporal Datascience


```{r setup, message=FALSE, warning=FALSE}
# Install missing packages if necessary
if (!requireNamespace("ranger", quietly = TRUE)) {
  install.packages("ranger", repos = "http://cran.us.r-project.org")
}

# Load required libraries
library(sf)
library(dplyr)
library(caret)
library(ranger)

# Load context layers
gpkg_osm <- "../data/osm.gpkg"
highway <- read_sf(gpkg_osm, "highway")
railway <- read_sf(gpkg_osm, "railway")

# Load trajectory dataset
tracks <- read_sf("../data/tracks_1.gpkg", layer = "training")
```



## 2. MOVEMENT FEATURE ENGINEERING

```{r}
# Prepare numeric coordinates for faster computation
coords <- st_coordinates(tracks)
tracks$X <- coords[,1]
tracks$Y <- coords[,2]

# Calculate basic movement parameters
tracks_features <- tracks |>
  arrange(user_id, datetime) |>
  group_by(user_id) |>
  mutate(
    lead_X = lead(X),
    lead_Y = lead(Y),
    time_diff = as.numeric(difftime(lead(datetime), datetime, units = "secs")),
    distance = sqrt((lead_X - X)^2 + (lead_Y - Y)^2),
    speed = ifelse(time_diff > 0, distance / time_diff, NA_real_),
    acceleration = (lead(speed) - speed) / time_diff,
    has_next = !is.na(lead_X) & !is.na(lead_Y) & time_diff > 0
  ) |>
  ungroup()

# Filter out invalid observations
tracks_features <- tracks_features |>
  filter(!is.na(speed) & is.finite(speed))
```

## 3. SPATIAL CONTEXT INTEGRATION

```{r}
set.seed(42)
# Sample points to balance performance and model robustness
tracks_context <- tracks_features |>
  slice_sample(n = min(50000, nrow(tracks_features)))

# Compute distance and attributes for nearest highway
nearest_hw_idx <- st_nearest_feature(tracks_context, highway)
tracks_context$distance_to_highway <- as.numeric(st_distance(tracks_context, highway[nearest_hw_idx, ], by_element = TRUE))
tracks_context$highway_type <- highway$highway[nearest_hw_idx]
tracks_context$cycleway <- highway$cycleway[nearest_hw_idx]

# Compute distance to nearest railway
nearest_rail_idx <- st_nearest_feature(tracks_context, railway)
tracks_context$distance_to_railway <- as.numeric(st_distance(tracks_context, railway[nearest_rail_idx, ], by_element = TRUE))
```

## 4. CUSTOM FEATURE ENGINEERING

```{r}
# Add binary indicator for railway proximity (< 50m)
tracks_context$close_to_railway <- tracks_context$distance_to_railway < 50

# Add binary indicator for major roads
major_road_types <- c("motorway", "trunk", "primary", "secondary")
tracks_context$is_major_road <- tracks_context$highway_type %in% major_road_types
```


## 5. DATA CLEANING & PREPARATION FOR ML

```{r}
# Handle missing values and convert to factors
ml_ready_data <- tracks_context |>
  mutate(
    highway = as.factor(highway_type),
    cycleway = as.factor(ifelse(is.na(cycleway), FALSE, cycleway)),
    close_to_railway = as.factor(close_to_railway),
    is_major_road = as.factor(is_major_road),
    mode = as.factor(mode)
  ) |>
  st_drop_geometry() |>
  na.omit()
```

## 6. MODEL TRAINING & EVALUATION

```{r}
# Define feature sets
movement_features <- c("speed", "acceleration")
context_features <- c("speed", "acceleration", "highway", "cycleway", 
                      "distance_to_railway", "distance_to_highway", 
                      "close_to_railway", "is_major_road")

# Split data into training (80%) and testing (20%)
set.seed(42)
train_indices <- createDataPartition(ml_ready_data$mode, p = 0.8, list = FALSE)
train_set <- ml_ready_data[train_indices, ]
test_set <- ml_ready_data[-train_indices, ]

# Define training formulas
formula_m1 <- as.formula(paste("mode ~", paste(movement_features, collapse = " + ")))
formula_m2 <- as.formula(paste("mode ~", paste(context_features, collapse = " + ")))

# Train Random Forest models
model_movement <- ranger(formula_m1, data = train_set, importance = "impurity", num.trees = 500)
model_context <- ranger(formula_m2, data = train_set, importance = "impurity", num.trees = 500, mtry = 3, min.node.size = 5)

# Evaluate models on test set
predictions_m1 <- predict(model_movement, data = test_set)$predictions
predictions_m2 <- predict(model_context, data = test_set)$predictions

confusion_matrix_m1 <- confusionMatrix(predictions_m1, test_set$mode)
confusion_matrix_m2 <- confusionMatrix(predictions_m2, test_set$mode)

# Output accuracy metrics
cat("--- Baseline Model Accuracy (Movement Only) ---\n")
print(confusion_matrix_m1$overall["Accuracy"])

cat("--- Enhanced Model Accuracy (Movement + Context) ---\n")
print(confusion_matrix_m2$overall["Accuracy"])
```

## Discussion of Results

The results show that including environmental context improves the prediction quite well. The baseline model, only using speed and acceleration, reaached an accuracy of 52.2%. This makes sense because these parameters can overlap in different transport mehtods. The Model with more context improved the overall accuracy to 70.7%. By adding features like distance to railway and distance to highway the model got some spatial context. Grouping the roads into major roads helped the algorithm to give slower or faster objects the context of why they move like this. The close to railway flag                        acted as a filter to eliminate background noise from more distent tracks. 

Working with this large dataset provided some insight on how to handle spatiotemporal data efficiently in R. Calculations like `st_nearest_feature` and `st_distance` are extremely heavy on the CPU when calculating this large dataset. So to make the workflow managable, a slicing strategy was implemented which sample the tracking points down to a smaller subset. This was done on the cost of accuracy in the model so it was important to find a optimum number of samples.

### Use of AI

AI tools (GitHub Copilot) were used in this exercise to assist with writing and debugging R code, as well as for proofreading and improving text.
